---
title: "Working With R Objects"
author: 'Jason Kinyua'
date: "6/19/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = TRUE)
```

## R Object/Data Types  

In R programming, the very basic data types are the R-objects. The simplest of these objects is the `atomic objects` which are six in number (covered in previous session). The other R-Objects are built upon these atomic vectors. 

### Derrived Object Types 

These are objects created from a combination of atomic types. 

#### Vectors  

When creating a vector object, function `c()` is used which means to combine the elements into a vector.  

```{r vectors}  
fruits <- c('grape', 'apple', 'pineaple', 'pawpaw','berry')

# show vector content
print(fruits)

# Get the class of the vector.
print(class(fruits))
```

```{r named-vectors}
# create a mixed named vector
sale <- c("color"="red", "item"="shirt", "size"="xL", "cost"=25, "currency"="$", "quantity"=5)

# show vector content
print(sale)

# Get the class of the vector.
print(class(sale))
```

##### Vector operations  

```{r c-gen}
# generate 1 to 10 using c()
values <- c(1:10)
values
```
```{r seq-gen}
# generate 1 to 20, increment by 2 using seq()
seq(from = 1, to = 20, by = 2)
```

```{r seq_len-gen}
# generate 20 values between 1 and 5
values <- seq(from = 1, to = 5, length.out = 20) # Note: 1 & 5 are now inclusive
print(values)
print(length(values))
```
```{r repeat-times}
# use rep() to repeat a vector n times
values <- c(1:5, 11:15, 21:25)
# will put 3 copies of the vector end-to-end
rep.values <- rep(values, times = 3)
print(values)
print(length(values))
print(rep.values)
print(length(rep.values))
```
```{r repeat-each}
# use rep() to repeat each element of vector n times
values <- c(1:5, 11:15, 21:25)
# repeats each element of the vector 3 times before moving on to the next.
rep.values <- rep(values, each = 3)
print(values)
print(length(values))
print(rep.values)
print(length(rep.values))
```
```{r arithmetic.operations}
values <- c(1:5, 11:15, 21:25)

sum(values)
```
```{r other-useful-math-functions}
# Statistics: sqrt(), mean(), median(), quantile(), min(), max(), sum(), cumsum(), prod(), cumprod()
# Trigonometric: sin(), cos(), tan(), acos(), asin(), atan() 
# Rounding: abs(), ceiling(), floor(), round(), sign(), signif(), trunc(), abs()
# Logarithms & Exponentials: exp(), log(), log10(), log2()
# 
values <- seq(from = 1, to = 5, length.out = 20)
round(values, 3)

# round(seq(from = 1, to = 5, length.out = 20), 3)
# Assign: median or random whole numbers, median of floating points (numeric)
```

```{r selecting-subsets}
x <- round(seq(1, 20, length.out=10), 2)
print(x)
print(x[5])
print(x[3:5])
print(x[7:9])
print(x[3:5]*x[7:9])
print(sum(x[3:5]*x[7:9]))
```
```{r select-named}
fruits <- c("berry","apple","grape","orange","peach","banana","pineaple","pawpaw")
sale <- c(2, 7, 3, 1, 0.8, 4, 5, 1.5)
# create named vector
names(sale) <- fruits
print(sale)
# subset
print(sale[c("banana", "grape", "orange")])
print(sort(sale))
```

#### Factors  

Factors are R’s way of storing categorical information. It stores the vector along with the distinct values of the elements in the vector as labels. The labels are always character irrespective of whether it is numeric or character or Boolean etc. in the input vector. They are useful in statistical modeling.  

Factors are created using the `factor()` function. The `nlevels` functions gives the count of levels.  
```{r create-a-factor}
# Create a vector.
apple_colors <- c('green','green','yellow','red','red','red','green')

# Create a factor object.
factor_apple <- factor(apple_colors)

# Print the factor.
print(factor_apple)
print(nlevels(factor_apple))
```

Classical use case of factors is preserving custom order of character objects which doesn't follow usual `sort` standard. Take for instance, ordering months of year.  

```{r sort-months}
year.months <- c("January","February","March","April","May","June","July","August","September","October","N```ovember","December")

my.months <- sample(year.months)
print("Randomized months")
my.months
print("Using sort to order months")
sort(my.months)
print("using factor to order months")
ordered <- factor(my.months, levels = year.months)
ordered
print("Desired months order")
levels(ordered)
```

#### Lists  

A list is an R-object which can contain many different types of elements inside it like vectors, functions and even another list inside it.

```{r using-lists}
list1 <- list(c(1:5), seq(1, 5, 3), rep(10, 3))
list2 <- list(c(7:10), seq(11, 15, 7), rep(21, 3))
final.list <- list(list1, list2)
print(list1)
print(list2)
print(final.list)
```
```{r named-list}
purchase <- list(
	"fruits" = list("orange"=5,"pineaple"=10,"banana"=3,"grape"=2,"berry"=4),
	"clothing" = list("shirt"=10,"trouser"=5,"coat"=3,"pant"=15)
	)
print(purchase)
```

```{r subset-list}
print(list1[[1]])
print(purchase$fruits)
```

##### List Operations  

```{r sale-costs}
# Assn
cost <- list() # create a list with cost of items in the purchase
# Use the purchase and cost lists to calculate amount payable by the customer, per item and total
```

#### Matrices  

A matrix is a two-dimensional rectangular data set. To create a matrix, first give `matrix` function an `atomic vector` to reorganize into a matrix. Then, define how many rows should be in the matrix by setting the `nrow` argument to a number. Alternatively, you can set the `ncol` argument, which tells R how many columns to include in the matrix. For sanity sake, using both `nrow` and `ncol` is recommended. A warning will be displayed if dimension condition is not met.

```{r a-matrix}
m <- matrix(round(seq(from=5, to = 200, length.out=35),3), nrow=7, ncol=5)
m
```

```{r matrix-order}
m <- matrix(round(seq(from=5, to = 200, length.out=35),3), nrow=7, ncol=5, byrow=TRUE)
m
```
```{r matrix-names}
matrix.labels <- list(c("a","b","c","d","e","f","g"), c("A","B","C","D","E"))
m <- matrix(round(seq(from=5, to = 200, length.out=35),3), nrow=7, ncol=5, byrow=TRUE, dimnames=matrix.labels)
m
```
```{r subset-matrix}
print("subset row")
m["a", ]
print("subset column")
m[,"A"]
print("subset value at [b, D]")
m["b","D"]
```

#### Arrays  

While matrices are confined to two dimensions, arrays can be of any number of dimensions. The `array` function takes a `dim` attribute which creates the required number of dimension. In the below example we create an array with 12 elements which are 2x2 matrices each.  

```{r create-array}
# Create an array.
a <- array(c(11:14, 21:24, 31:34), dim = c(2, 2, 3))
print(a)
```

#### Data Frames  

Data frames are tabular data objects. Unlike a matrix in data frame each column can contain different modes of data. The first column can be numeric while the second column can be character and third column can be logical. It is a list of vectors of equal length.  

Data Frames are created using the `data.frame()` function.  

```{r sample-df}
# Create the data frame.
bmi <- 	data.frame(
   gender = c("Male", "Male","Female"), 
   height = c(152, 171.5, 165), 
   weight = c(81,93, 78),
   Age = c(42,38,26)
)
print(bmi)
```

#### Others  

- date
- tibble
- data.table

##### Date  

```{r dates}
date <- as.Date("1/25/2021", format="%m/%d/%Y")
date
```

```{r date.ops}
date1 <- as.Date("1/25/2021", format="%m/%d/%Y")
date2 <- as.Date("4/1/2021", format="%m/%d/%Y")
duration <- seq(from = date1, to = date2, by = 1)
duration
```

```{r extract.date.parts}
format(date1, "%a %d %B, %Y")
```

```{r current.date}
date.now <- Sys.Date()
date.now
time.now <- Sys.time()
time.now
```

#### Useful Functions not covered  

There are several base R functions that are very useful that you might want to know nonetheless, such as `paste()` and `paste0()`  

Let's take a look at each of them:  

```{r extras}
paste("Hello", "PyR4DS")
```

you can change the separator if needed  

```{r}
paste("Hello", "PyR4DS", sep=" :) ")
```

`paste0()` is the same as `paste()` but does not have any `sep` argument neither does it separate words by a whitespace by default.  

```{r}
paste0("Hello", "PyR4DS")
paste0(c("Hello","PyR4DS"), collapse=" ")
```

To change the case of characters, you can use `toupper()` and `tolower()`:  

```{r}
greeting <- paste0(c("Hello","PyR4DS"), collapse=" ")
tolower(greeting)
toupper(greeting)
greeting
```

```{r logical.check}
ifelse(3 > 5, "3 is greater than 5", "3 is not greater than 5")
```